---
title: "Week 3 Strings and Regex"
author: "Ethan Haley"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(stringr)
```

#### Get some data about college majors, from fivethirtyeight.com's Github

```{r}
dataURL <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/college-majors/majors-list.csv"
majors <- read.csv(url(dataURL))
majors <- majors$Major
majors[1:11]
```

##  Identify the majors that contain either "DATA" or "STATISTICS"

```{r}
majors[str_detect(majors, "DATA|STATISTICS")]
```

### -----------------------------------------------

## Practice manipulating strings with regex in R

```{r}
fruitvec <- c("bell pepper", "bilberry", "blackberry", "blood orange",
              "blueberry", "cantaloupe", "chili pepper", "cloudberry",
              "elderberry", "lime", "lychee", "mulberry", "olive",
              "salal berry")
fruitvec
```

##### Now the task is to create a string representation of ```fruitvec```: \
#####  

   
```{r}
strfruitvec <- str_flatten(fruitvec, collapse = '", "')
cat(c('c("', strfruitvec, '")'), sep = '')
```
##### So the ```cat``` function produces the desired output format, but in order to produce an actual object that is a string and includes double quotes around each item, I don't know if that's possible, without backslashes appearing in the object.
#  

```{r}
stringy_vec <- function(charvec) {
  flat <- str_flatten(charvec, collapse = '", "')
  cat(c('c("', flat, '")'), sep = '')
}
cheeses <- c('gouda', 'brie', 'stilton', 'american', 'string cheese')
stringy_vec(cheeses)
```


### ----------------------------------------------------
## Describe, in words, what these expressions will match:
# 
##### ```(.)\1\1```
##### A character that occurs 3 times in a row
#
##### ```"(.)(.)\\2\\1"```
##### A palindromic series of 4 characters, like "anna" or "zzzz"
#
##### ```(..)\1```
##### Four characters where 3 and 4 are the same as 1 and 2, like "yoyo"
#
##### ```"(.).\\1.\\1"```
##### A 5-char sequence where chars 1, 3, and 5 are the same, like "orono"
#
##### ```"(.)(.)(.).*\\3\\2\\1"```
##### A series of at least 6 chars, ending with the same 3 it started with,
##### but in reversed order, like "redder" or "madam, i'm adam"

### --------------------------------------------------

## Construct regular expressions to match words that:
#
##### --Start and end with the same character.
##### ```  "\\b(.)\\S*\\1\\b"``` allows internal numbers and symbols, so it 
##### matches "pop-up" and "s#$%s", e.g.
#
##### --Contain a repeated pair of letters (e.g. "church" contains "ch" repeated twice.)
##### ``` "\\b.*([a-zA-Z][a-zA-Z])[a-zA-Z-]*\\1[a-zA-Z]*\\b" ``` allows only letters,
##### plus hyphens between the repeated pairs, so it matches 'mai-tai' 
##### but only the "yoyo" part of "yoyo-lover", e.g.
#
##### --Contain one letter repeated in at least three places (e.g. "eleven" contains three "e"s.)
##### ``` "\\b[a-zA-Z]*([a-zA-Z])[a-zA-Z]+\\1[a-zA-Z]+\\1[a-zA-Z]*\\b" ```




...

