---
title: "Week 1: Loading and Subsetting Data"
author: "Ethan Haley"
date: "2/5/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Overview of [FiveThirtyEight's article exploring why people don't vote](https://projects.fivethirtyeight.com/non-voters-poll-2020-election/)

This article, using polling data gathered 49-39 days before the latest U.S. presidential election, in which the highest percentage of eligible voters since 1900 actually voted, takes a look at why non-voters may choose or need to remain on the sidelines.  Presumably the conclusions drawn from this type of study could be used to help create policies that lessen the pull of whatever it is that's holding these people back, but the darker undercurrent is the fact that there's a significant proportion of policy-makers interested in exacerbating these problems.

The data come from a carefully selected group of voting age Americans with proven voting histories, chosen and weighted to approximate the general population.  Also some respondents with no voting histories who expressly stated not voting in the past are included, to represent a typical mix of voters and non-voters.  Here's how this group's track record breaks down:

```{r}

dataUrl <- "https://raw.githubusercontent.com/ebhtra/msds-607/main/wk1_R_538/non-voters/nonvoters_data.csv"

noVotes <- read.csv(url(dataUrl))
# rearrange the categories in their natural order
noVotes$voter_category <- factor(noVotes$voter_category, levels = c('rarely/never', 'sporadic', 'always'))
barplot(table(noVotes$voter_category), main = "Voting History of Respondents", ylab = "Count of People")
```

If the largest group, who votes sporadically, divides somewhat evenly between the other two groups in any given year, this is a typical group of people, since about 50-60% of eligible voters have voted in each general election since 1900.


### When people don't vote, what are their reasons?

```{r}
# leave out most of the demographic info and put names on the rest
# based on the "nonvoters_codebook.pdf" that accompanies these data
oldcols <- c("Q2_1", "Q2_2", "Q2_3", "Q2_4", "Q2_5", "Q2_6", "Q2_7", "Q2_8", "Q2_9", "Q2_10", "Q5", "Q8_1", "Q8_2", "Q8_3", "Q8_4", "Q8_5", "Q8_6", "Q8_7", "Q8_8", "Q8_9", "Q17_1", "Q17_2", "Q17_3", "Q17_4", "Q18_1", "Q18_2", "Q18_3", "Q18_4", "Q18_5", "Q18_6", "Q18_7", "Q18_8", "Q18_9", "Q18_10", "Q20", "Q21", "Q22", "Q26", "Q29_1", "Q29_2", "Q29_3", "Q29_4", "Q29_5", "Q29_6", "Q29_7", "Q29_8", "Q29_9", "Q29_10", "Q30", "voter_category")
newcols <- c("good_Am_vote", "good_Am_jury", "good_Am_politics", "good_Am_flag", "good_Am_census", "good_Am_Pledge", "good_Am_military", "good_Am_respect_opp", "good_Am_God", "good_Am_protest", "winner_matters", "trust_prez", "trust_congr", "trust_court", "trust_CDC", "trust_elec_offic", "trust_FBI", "trust_media", "trust_police", "trust_postal", "trust_machine", "trust_paper", "trust_mail", "trust_e-vote", "problem_ID", "prob_find_poll", "prob_deadline", "prob_access", "prob_ballot_help", "prob_provisional", "prob_work", "prob_long_line", "prob_name_list", "prob_absentee", "registered", "plan_to_vote", "why_not_voting", "voting_history", "disliked_cands", "geoloc_nullified", "no_effect_on_me", "broken_system", "tried_but_failed", "unsure_eligible", "irrelevant_issues", "all_cand_same", "opposed_to_voting", "other", "party", "voter_category")

noVotes <- noVotes[oldcols]
colnames(noVotes) <- newcols

# convert 1=yes and 2=no into booleans 1=T and 2=F, for appropriate columns
yes_no <- c("winner_matters", "problem_ID", "prob_find_poll", "prob_deadline", "prob_access", "prob_ballot_help", "prob_provisional", "prob_work", "prob_long_line", "prob_name_list", "prob_absentee", "registered")
noVotes[yes_no] <- lapply(noVotes[yes_no], function(x) as.logical(abs(x-2)))
# convert plan_to_vote to factor, discarding the unanswereds (-1's)
noVotes$plan_to_vote <- factor(noVotes$plan_to_vote, levels = c(1:3), labels = c("Yes", "No", "Maybe"))
# factorize party affiliation col
noVotes$party <- factor(noVotes$party, levels = c(1:5), labels = c("Rep", "Dem", "Ind", "Other", "NoPref"))
# factorize reasons for not planning to vote
noVotes$why_not_voting <- factor(noVotes$why_not_voting, levels = c(1:7), labels = c("no_time", "feel_abandoned", "how_register", "dont_want_register", "ineligible", "doesnt_matter", "other"))
# long labels, so go horiz, and increase left margin
par(mar = c(10, 10, 3, 5))
barplot(sort(table(noVotes$why_not_voting)), horiz = T, las = 1, main = "Why Some People Are Planning Not to Vote", xlab = "Number of People")
# plot problems voting in past
colsums <- colSums(noVotes[c("problem_ID","prob_find_poll", "prob_deadline", "prob_access", "prob_ballot_help", "prob_provisional", "prob_work", "prob_long_line", "prob_name_list", "prob_absentee")])
barplot(sort(colsums), horiz = T, las = 1, main = "Problems People or Household Members Had Voting in Past")

```

### Are they planning to vote in the upcoming election?
```{r}
table(noVotes$plan_to_vote)
```

Looks like this group is fired up to vote, although it seems likely a lot of them didn't end up following through.  Speaking of what fires these people up, the survey includes some interesting questions about what makes someone be a good American.  Each trait is rated 1-4 from very important to unimportant, and this gives us a chance to get a quick look at where people differ on things they feel strongly about, just by looking at the variance in answers within each quality.

```{r}
vars <- c(var(noVotes$good_Am_flag[noVotes$good_Am_flag > 0]), var(noVotes$good_Am_census[noVotes$good_Am_census > 0]), var(noVotes$good_Am_Pledge[noVotes$good_Am_Pledge > 0]), var(noVotes$good_Am_God[noVotes$good_Am_God > 0]), 
var(noVotes$good_Am_vote[noVotes$good_Am_vote > 0]), 
var(noVotes$good_Am_jury[noVotes$good_Am_jury > 0]), var(noVotes$good_Am_military[noVotes$good_Am_military > 0]), var(noVotes$good_Am_politics[noVotes$good_Am_politics > 0]) )

ord <- order(vars)

labls <- c('display flag', 'be in census', 'know Pledge', 'believe in God', 'vote', 'serve as juror', 'support military', 'follow politics')

par(mar = c(10, 10, 3, 5))
barplot(vars[ord], names.arg = labls[ord], horiz = T, las = 1, xlab = 'Variance of Opinion', main = 'Differences in Opinions About What Makes a Good American')
```

### Finally, a brief look at how much people trust various institutions, according to respondents here.

Again these are rated 1-4, from most trust to least for each institution, so we could see how much people's opinions differed again, but let's instead just see how the overall levels of trust measure up. I'll calculate (4 - rating) for each one, so that the highest trust is 3 and the lowest is 0.

```{r}
trust <- c(mean(noVotes$trust_prez[noVotes$trust_prez > 0]), mean(noVotes$trust_congr[noVotes$trust_congr > 0]), mean(noVotes$trust_court[noVotes$trust_court > 0]), mean(noVotes$trust_CDC[noVotes$trust_CDC > 0]), mean(noVotes$trust_elec_offic[noVotes$trust_elec_offic > 0]), mean(noVotes$trust_FBI[noVotes$trust_FBI > 0]), mean(noVotes$trust_media[noVotes$trust_media > 0]), mean(noVotes$trust_police[noVotes$trust_police > 0]), mean(noVotes$trust_postal[noVotes$trust_postal > 0]))
# subtract means from 4 to make it more intuitive (vs. plotting distrust)
trust <- 4 - trust
ord <- order(trust)

labls <- c('president', 'congress', 'supreme court', 'C.D.C.', 'election officials', 'FBI/CIA', 'media', 'police', 'post office')

par(mar = c(10, 10, 3, 5))
barplot(trust[ord], names.arg = labls[ord], horiz = T, las = 1, xlab = 'Amount of Trust', main = "People's Average Trust in Entities")
```


### Possible Next Steps

I chose to remove the demographic elements of this survey, and instead just treat people as people, but obviously there are insights to be gained by stratifying people by gender, race, age, income, political party, etc.  It certainly felt nice not to, in this short study, but the data are all there, whenever someone wants to take a more "nuanced" look at the correlations involved, as the authors at FiveThirtyEight did.


========================================================================================



